\bf Hierarchical Network Structure Promotes Dynamical Robustness

### 2014-11-26 tex/abstract-cited.tex
minor edits

%!TEX root = ../paper.tex
\begin{abstract}
The analysis of dynamical systems that attempts to model chemical reaction, gene-regulatory, population, and ecosystem networks all rely on models having interacting components~\cite{RossCr2003,Alon2006,Palsson2006,HamidBolouri2008,Palsson2011a,Voit2012,Sauro2012}. When the details of these interactions are unknown for biological systems of interest, one effective approach is to study the dynamical properties of an ensemble of models determined by evolutionary constraints that may apply to all such systems~\cite{Gardner1970,May1972,Cohen1984,May1972a,Radius2014}. One such constraint is that of dynamical robustness, the probability of a stable system to remain stable under perturbations to the strength of interactions among its components~\cite{WADDINGTON1942a,Wagner1997,Rutherford1998,VanNimwegen1999,Siegal2002,Bergman2003,Ciliberti2007b,Ciliberti2007,Draghi2010,Wagner2013}. Despite previous investigations, the relationship between dynamical robustness---an important functional characteristic of many biological systems---and network structure is poorly understood. Here we analyze the stability and robustness of a large class of dynamical systems and demonstrate that the most hierarchical network structures are the most robust. In particular, we determine the probability distribution of robustness over system connectivity and show that robustness is maximized by maximizing the number of links between strongly connected components of the graph representing the underlying system connectivity. We demonstrate that this can be understood in terms of the fact that permutation of strongly connected components is a fundamental symmetry of dynamical robustness, which applies to networks of any number of components and is independent of the distribution from which the strengths of interconnection among components are sampled.  The classification of dynamical robustness based upon a purely topological property provides a fundamental organizing principle that can be used in the context of experimental validation to select among models that break or preserve network hierarchy. This result contributes to an explanation for the observation of hierarchical modularity in biological networks at all scales~\cite{Zhao2006,Ravasz2002,Bhardwaj2010,Colm,Corominas-Murtra2013}.
\end{abstract}

>>>>

%!TEX root = ../paper.tex
\begin{abstract}
The analysis of dynamical systems that attempts to model chemical reaction, gene-regulatory, population, and ecosystem networks all rely on models having interacting components~\cite{RossCr2003,Alon2006,Palsson2006,HamidBolouri2008,Palsson2011a,Voit2012,Sauro2012}. When the details of these interactions are unknown for biological systems of interest, one effective approach is to study the dynamical properties of an ensemble of models determined by evolutionary constraints that may apply to all such systems~\cite{Gardner1970,May1972,Cohen1984,May1972a,Radius2014}. One such constraint is that of dynamical robustness, the probability of a stable system to remain stable under perturbations to the strength of interactions among its components~\cite{WADDINGTON1942a,Wagner1997,Rutherford1998,VanNimwegen1999,Siegal2002,Bergman2003,Ciliberti2007b,Ciliberti2007,Draghi2010,Wagner2013}. Despite previous investigations, the relationship between dynamical robustness---an important functional characteristic of many biological systems---and network structure is poorly understood. Here we analyze the stability and robustness of a large class of dynamical systems and demonstrate that the most hierarchical network structures, those equivalent to the total ordering, are the most robust. In particular, we determine the probability distribution of robustness over system connectivity and show that robustness is maximized by maximizing the number of links between strongly connected components of the graph representing the underlying system connectivity. We demonstrate that this can be understood in terms of the fact that permutation of strongly connected components is a fundamental symmetry of dynamical robustness, which applies to networks of any number of components and is independent of the distribution from which the strengths of interconnection among components are sampled.  The classification of dynamical robustness based upon a purely topological property provides a fundamental organizing principle that can be used in the context of experimental validation to select among models that break or preserve network hierarchy. This result contributes to an explanation for the observation of hierarchical modularity in biological networks at all scales~\cite{Zhao2006,Ravasz2002,Bhardwaj2010,Colm,Corominas-Murtra2013}.
\end{abstract}

### 2014-11-26 tex/introduction.tex
minor edits

Here we determine the relationship between network hierarchy, a topological property, and the probability of \emph{robustness}, a dynamical property. Robustness is of interest to biological systems at all scales, and has been previously studied in the context of gene-regulatory networks \cite{WADDINGTON1942a,VanNimwegen1999,Siegal2002,Ciliberti2007b,Ciliberti2007,Wagner2013}. Intuitively, robustness is the probability of stability to perturbation in the system structure or its parameters for systems which have already been determined to be stable in the sense of linear stability analysis \cite{Davis1962}. We demonstrate that systems with the most hierarchical network topology exhibit maximal robustness and explain why this results from an invariance of robustness to particular kinds of transformations of the network topology.

>>>>

Here we determine the relationship between network hierarchy, a topological property, and the probability of \emph{robustness}, a dynamical property. Robustness is of interest to biological systems at all scales, and has been previously studied in the context of gene-regulatory networks \cite{WADDINGTON1942a,VanNimwegen1999,Siegal2002,Ciliberti2007b,Ciliberti2007,Wagner2013}. Intuitively, robustness is the probability of stability to potentially large perturbations of the parameters for systems which have already been determined to be stable in the sense of linear stability analysis \cite{Davis1962}. The maximally hierarchical network is considered to be the graph associated to the total ordering, which is the maximally connected directed acyclic graph for a given system size. Here, hierarchy is then defined as the edit distance from this maximally hierarchical network. We demonstrate that systems with the most hierarchical network topology exhibit maximal robustness and explain why this results from an invariance of robustness to particular kinds of transformations of the network topology. Moreover these results hold for networks of arbitrary size and independent of the probability distribution from which the strengths of interaction are sampled.
